---
title: "Assignment 3 DATA 607"
author: "Barakat Adigun"
date: "2025-02-18"
output:
  html_document: default
  pdf_document: default
---

```{r}
library(tidyverse)
```

## Instructions
Please deliver links to an R Markdown file (in GitHub and rpubs.com) with solutions to the problems below.  You may work in a small group, but please submit separately with names of all group participants in your submission.

## Normalization Question 1
Provide an example of at least three dataframes in R that demonstrate normalization. The dataframes can contain any data, either real or synthetic. Although normalization is typically done in SQL and relational databases, you are expected to show this example in R, as it is our main work environment in this course.

DataFrame 1:Product ratings
```{r}
## Product ratings
Product_ratings <- data.frame(
  ProductID = 1:6,
  Rating = c(4.1, 3.7, 5.0, 4.4, 3.2, 4.5),
  Price = c(50, 60, 80, 70, 40, 45)
)

print(Product_ratings)
```

Example 1 using Dataframe 1: Min-Max Normalization
Min-max normalization scales numerical data to a fixed range, typically between 0 and 1. The benefits of using this strategy includes; Consistent Scale Across Variables which ensures all features (such as, rating, pricing) are on the same scale, preventing variables with larger ranges from dominating. It also Preserved Relationships: The relative relationships between data points are maintained, ensuring data integrity.
```{r}
# Function to normalize each numeric column in a dataframe
normalize_df <- function(df) {
  # Loop through all numeric columns and apply Min-Max normalization
  df_normalized <- df
  for (col in names(df)) {
    if (is.numeric(df[[col]])) {
      min_val <- min(df[[col]], na.rm = TRUE)
      max_val <- max(df[[col]], na.rm = TRUE)
      df_normalized[[col]] <- (df[[col]] - min_val) / (max_val - min_val)
    }
  }
  return(df_normalized)
}

Product_ratings_normalized <- normalize_df(Product_ratings)

print(Product_ratings_normalized)
```

DataFrame 2: Sales data with different units
```{r}
## Sales data with different units
Sales_Data <- data.frame(
  SalesID = 1:5,
  UnitsSold = c(120, 150, 80, 200, 50),
  Revenue = c(1000, 2000, 1500, 2500, 1200)
)

print(Sales_Data)
```

Example 2 using Dataframe 2: Normalization Using a Custom Range
Custom range normalization allows you to scale data to any desired range, such as between -1 and 1, or any other interval. By applying a custom range, you can adjust the normalized data to a more specific scale that better suits the problem one is working on.
```{r}
# Custom normalization (scale to range between -1 and 1)
Sales_Data_custom_normalized <- Sales_Data %>%
  mutate(
    Revenue_custom_normalized = 2 * ((Revenue - min(Revenue)) / (max(Revenue) - min(Revenue))) - 1
  )

# View the original and custom normalized data
print(Sales_Data_custom_normalized)
```

DataFrame3: Student Data including Student Test Scores and Study Hours
```{r}
# DataFrame: Student Test Scores and Study Hours
Student_Data <- data.frame(
  StudentID = 1:6,
  TestScore = c(78, 85, 92, 65, 88, 73),
  StudyHours = c(10, 15, 12, 8, 20, 11)
)

print(Student_Data)
```

Example 3 using Dataframe 3: Z-Score Normalization (Standardization)
Z-score normalization transforms the data so that the mean is 0 and the standard deviation is 1. This method is particularly useful when the data has varying scales or distributions. It helps improve model performance, ensures fair contribution from all features, and speeds up optimization processes.
```{r}
# Z-Score normalization for Height and Weight
Student_Data_standardized <- Student_Data %>%
  mutate(
    TestScore_standardized = (TestScore - mean(TestScore)) / sd(TestScore),
    StudyHours_standardized = (StudyHours - mean(StudyHours)) / sd(StudyHours)
  )

# View the original and standardized data
Student_Data_standardized
```

## Character Manipulation Question 2
Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r}
# Load library
library(dplyr)
```

```{r}
# Load the dataset 
url <- "https://raw.githubusercontent.com/Badigun/Data-607-Assignments/refs/heads/main/majors-list.csv"
majors_list <- read.csv(url)

# Check the dataset
head(majors_list)

# Filter majors that contain "DATA" or "STATISTICS"
filtered_Majors_Data <- majors_list %>%
  filter(grepl("DATA|STATISTICS", Major, ignore.case = TRUE))

# Display the filtered majors
print(filtered_Majors_Data$Major)
```

## The two exercises below are taken from R for Data Science, 14.3.5.1 in the on-line version:

## Question 3: Describe, in words, what these expressions will match:

(.)\1\1
Answer: This expression matches any three consecutive characters where the first character is repeated two more times. For example, it will match strings like "aaa", "bbb" where the first character repeats twice more.

"(.)(.)\\2\\1"
Answer: This expression matches a string with two characters, followed by those same two characters in reverse order. For example, it will match strings like "abba". The first character is matched again as the last character, and the second character is matched again as the second-to-last character.

(..)\1
Answer: This expression matches a string where the first two characters are exactly repeated after them. For example, it will match strings like "abab", or "1212". It requires that the first two characters are the same as the next two characters.

"(.).\\1.\\1"
Answer: This expression matches strings where the first character appears three times, with two other characters in between. For example, it will match strings like "aXa".

"(.)(.)(.).*\\3\\2\\1"
Answer: This expression matches strings where the first three characters are followed by any characters, and then those three characters are repeated in reverse order. For example, it will match strings like "abcxyzcba", or "123xyzyx321". The first three characters appear again after some characters, but in reverse order.

## Question 4: Construct regular expressions to match words that:
- Start and end with the same character - \\b([a-zA-Z]).*\\1\\b
```{r}
# Word examples
word <- c("plump", "pen", "radar", "noon", "civic", "world")

# Regex pattern to match words that start and end with the same character
pattern <- "\\b([a-zA-Z]).*\\1\\b"

# use grep to find words that match the pattern
matches <- grep(pattern, word, value = TRUE)
print(matches)
```

- Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.) - \\b([a-zA-Z]{2}).*\\1\\b
```{r}
# Sample text
text <- c("church", "money", "success", "monkey", "love")

pattern <- "\\b([a-zA-Z]{2}).*\\1\\b"

# Find words that match the pattern
matches <- grep(pattern, text, value = TRUE)
print(matches)
```

- Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.) - \\b([a-zA-Z]).*\\1.*\\1\\b
```{r}
text <- c("love", "balloon", "success", "hello", "apple", "test")

pattern <- "\\b([a-zA-Z]).*\\1.*\\1\\b"

matches <- grep(pattern, text, value = TRUE)
print(matches)
```



____