---
title: "Assignment 4 DATA 607 Tidying and Transforming Data"
author: "Barakat Adigun"
date: "2025-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The chart above describes arrival delays for two airlines across five destinations. Your task is to:

## (1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above. You’re encouraged to use a “wide” structure similar to how the information appears above, so that you can practice tidying and transformations as described below.
Answer: The .CSV file created will be attached to the assignment file

## (2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and transform your data.

```{r}
library(tidyverse)
```

```{r}
library(dplyr)
```

```{r}
## Load Data into R

Flight_ArrDelay <- read.csv(file="https://raw.githubusercontent.com/Badigun/Data-607-Assignments/refs/heads/main/Flights%20Arrival%3Adelays%20DATA%20607.csv", header = TRUE, sep = ",")
```

```{r}
## View the loaded data to make sure it's loaded correctly
head (Flight_ArrDelay)
```

```{r}
# Create the dataset Manually
FlightArrDelay <- tibble(
  Destination = c("Los Angeles", "Phoenix", "San Diego", "San Francisco", "Seattle"),
  `Alaska (On Time)` = c(497, 221, 212, 503, 1841),
  `Alaska (Delayed)` = c(62, 12, 20, 102, 305),
  `AM West (On Time)` = c(694, 4840, 383, 320, 201),
  `AM West (Delayed)` = c(117, 415, 65, 129, 61)
)

# View the data
head(FlightArrDelay)
```

```{r}
# (tidyr) Pivot the data to a long format for easier analysis
longer_FlightArrDelay <- FlightArrDelay %>%
  pivot_longer(
    cols = c(`Alaska (On Time)`, `Alaska (Delayed)`, `AM West (On Time)`, `AM West (Delayed)`),  
    names_to = c("Airline", "Status"),                     
    names_pattern = "(.*) \\((.*)\\)",                    
    values_to = "Flights"
  )

# View the tidied data
head(longer_FlightArrDelay)
```

```{r}
# Filter the data for "On Time" flights
ontime_flights <- longer_FlightArrDelay %>%
  filter(Status == "On Time")

# View the filtered data
head(ontime_flights)
```

```{r}
# Filter the data for "Delayed" flights
Delayed_flights <- longer_FlightArrDelay %>%
  filter(Status == "Delayed")

# View the filtered data
head(Delayed_flights)
```

```{r}
# View the total flights for each airline and status
totalflights <- longer_FlightArrDelay %>%
  group_by(Airline, Status) %>%
  summarise(Total_Flights = sum(Flights)) %>%
  ungroup()

print(totalflights)
```

## (3) Perform analysis to compare the arrival delays for the two airlines.

```{r}
# Calculate the proportion of delayed flights for each airline
proportion_delayed <- longer_FlightArrDelay %>%
  group_by(Airline) %>%
  summarise(
    Total_On_Time = sum(Flights[Status == "On Time"]),
    Total_Delayed = sum(Flights[Status == "Delayed"]),
    Total_Flights = sum(Flights),
    Proportion_Delayed = Total_Delayed / Total_Flights
  ) %>%
  ungroup()

print(proportion_delayed)
```

```{r}
# Calculate the average number of delayed flights for each airline
avg_delayed_flights <- longer_FlightArrDelay %>%
  filter(Status == "Delayed") %>%
  group_by(Airline) %>%
  summarise(Average_Delayed_Flights = mean(Flights)) %>%
  ungroup()

print(avg_delayed_flights)
```

## Conclusion

The analysis of arrival delays for Alaska Airlines and AM West Airlines revealed the following:

- Proportion of Delayed Flights: Alaska Airlines has a higher proportion of delayed flights (13.3%) compared to AM West Airlines (10.9%).

- Average Number of Delayed Flights: AM West Airlines experiences, on average, more delayed flights (157.4 per destination) than Alaska Airlines (100.2 per destination).


____