---
title: "Assignment 6 DATA 607 WEEK 9"
author: "Barakat Adigun"
date: "2025-04-08"
output:
  html_document: default
  pdf_document: default
---


## Assignment – Web APIs

The New York Times web site provides a rich set of APIs, as described here: https://developer.nytimes.com/apis

You’ll need to start by signing up for an API key.

Your task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and transform it into an R DataFrame.

## Introduction:
For this assignment, I will be using the NYT Books API, which is great for exploring bestseller lists or get reviews of books from The New York Times.

```{r}
# Load Libraries

library(httr)
library(jsonlite)
library(dplyr)
```

```{r}
# API URL to be used
# https://api.nytimes.com/svc/books/v3/lists/current/hardcover-fiction.json?api-key=YOUR_API_KEY

# API APP NAME
# BOOKS_API
```

```{r}
usethis::edit_r_environ()
```

```{r}
# This function is to get the NYT Best Sellers list
getting_nyt_bestsellers <- function(list_name = "hardcover-fiction", api_key) {
  # Construct API URL
  base_url <- "https://api.nytimes.com/svc/books/v3/lists/current"
  full_url <- paste0(base_url, "/", list_name, ".json?api-key=", api_key)
  
  # Make API request
  response <- GET(full_url)
  
  # Checking for errors
  if (http_error(response)) {
    stop("Request failed: ", status_code(response))
  }
  
  # Parse JSON response
  content_json <- content(response, as = "text")
  parsed <- fromJSON(content_json, flatten = TRUE)
  
  # Extract books and clean into a DataFrame
  books_df <- parsed$results$books %>%
    select(rank, title, author, description, publisher, weeks_on_list, amazon_product_url)
  
  return(books_df)
}

# Example usage
api_key <- Sys.getenv("NYT_API_KEY")

# Get current bestsellers in hardcover fiction
bestsellers_nyt <- getting_nyt_bestsellers("hardcover-fiction", api_key)

# View result
head(bestsellers_nyt)
```

## In Addition, I will like to analysis the data a bit

# This is to show a basic summary statics of the current bestsellers in hardcover fiction
```{r}
# How many books are on the list?
nrow(bestsellers_nyt)

# Summary of weeks on the list
summary(bestsellers_nyt$weeks_on_list)
```

# Top authors with more books on the hardcover fiction
```{r}
# Count authors with multiple books on the list
bestsellers_nyt %>%
  group_by(author) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```

# Visualization: Weeks on List
```{r}
library(ggplot2)

ggplot(bestsellers_nyt, aes(x = reorder(title, weeks_on_list), y = weeks_on_list)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "NYT Hardcover Fiction Bestsellers",
       x = "Book Title",
       y = "Weeks on List")
```

## Conclusion: Web APIs; NYT Books API; NYT Best Sellers Analysis

I was able to successfully retrieved data from the NYT Books API and analyzed the hardcover fiction bestseller list. 

Key steps included:

Data Retrieval: Pulled current bestsellers and transformed the JSON data into an R data.frame.

Summary Stats: We explored the number of books and their weeks on the list.

Top Authors: Identified authors with the most books on the list.

Visualizations: Created a bar chart of books’ weeks on the list.

This analysis provides insights into bestseller trends and author popularity.


_________________